Comparison of Terrain Following and Cut Cell Grids with a Non-Hydrostatic Model

### 2015-04-28 msc-publication.tex
more corrections from John, swap Klemp2011 fig2 for fig4

The results of the tracer advection test show that, given an advection scheme of a sufficiently high order, distortions in the grid do not significantly distort the tracer.

@@

At $t = \SI{10000}{\second}$ the tracer has passed over the mountain.  The horizontal position of the tracer centre can be calculated by integrating along the trajectory to find $t$, the time taken to pass from one side of the mountain to the other:
\begin{multline}
t = \frac{x}{u_0} - \frac{h_0}{16 u_0 H} \left[ 4x + \frac{\sin 2 (\alpha + \beta) x}{\alpha + \beta} \right.+ \\
\left. \frac{\sin 2(\alpha - \beta) x}{\alpha - \beta} + 2 \left( \frac{\sin 2\alpha x}{\alpha} + \frac{\sin 2\beta x}{\beta} \right) \right]
\end{multline}
Hence, we find that \(x(t=\SI{10000}{\second}) = \SI{51577.4}{\meter}\).  Because the velocity field is non-divergent, the flow accelerates over mountain ridges and the tracer travels \SI{1577.4}{\meter} further compared to advection in the purely horizontal velocity field.  Tracer height is unchanged downwind of the mountains because advection is along the terrain following coordinate surface.

\begin{table*}[t]
	\caption{Minimum and maximum tracer magnitudes and \(\ell^2\) error norms at \(t = \SI{10000}{\second}\) in the horizontal and terrain following tracer advection tests using centred linear and cubic upwind-biased schemes.}
\label{tab:advection}
%
\centering
\footnotesize
\begin{tabular}{ r @{\hspace{0.5em}} l l l @{\hspace{0.2em}} l l l @{\hspace{0.5em}} l l l @{\hspace{0.2em}} l l l}
\hline\hline
& \multicolumn{6}{c}{Horizontal advection} & \multicolumn{6}{c}{Terrain following advection} \\
& \multicolumn{3}{c}{Centred linear} & \multicolumn{3}{c}{Cubic upwind-biased} & \multicolumn{3}{c}{Centred linear} & \multicolumn{3}{c}{Cubic upwind-biased} \\
& \(\ell^2\) error & min & max & \(\ell^2\) error & min & max & \(\ell^2\) error & min & max & \(\ell^2\) error & min & max \\
\hline
Analytic  & 0 & 0 & 1 & 0 & 0 & 1 & 0 & 0 & 1 & 0 & 0 & 1 \\
BTF 	  & \num{0.0318} & \num{-0.442} & \num{0.849} & \num{0.00767} & \num{-0.0433} & \num{0.928} & \num{0.00251} & \num{-0.0244} & \num{0.985} & \num{0.00154} & \num{-0.0105} & \num{0.983} \\
SLEVE 	  & \num{0.00233} & \num{-0.0253} & \num{0.945} & \num{0.00108} & \num{-0.0106} & \num{0.981} & \num{0.0298} & \num{-0.258} & \num{0.865} & \num{0.0122} & \num{-0.0273} & \num{0.864} \\
Cut cell  & \num{0.00224} & \num{-0.0253} & \num{0.985} & \num{0.000579} & \num{-0.00670} & \num{0.982} & \num{0.0318} & \num{-0.270} & \num{0.846} & \num{0.0136} & \num{-0.0297} & \num{0.850} \\
No terrain & \num{0.00224} & \num{-0.0253} & \num{0.985} & \num{0.000577} & \num{-0.00670} & \num{0.982} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} \\
\hline
\end{tabular}
\end{table*}

@@

We conclude from this test that accuracy depends upon alignment of the flow with the grid, and accuracy is not significantly reduced by grid distortions.  Error on the BTF grid is comparable with the error on the SLEVE grid in the horizontal advection test.

@@

The initial thermodynamic conditions are in discrete hydrostatic balance, having a surface temperature of $\theta_0 = \SI{288}{\kelvin}$ and constant stability with Brunt-V\"ais\"al\"a frequency $N = \SI{0.01}{\per\second}$ everywhere, except for a more stable layer of $N = \SI{0.02}{\per\second}$ between $\SI{2}{\kilo\meter} \leq z \leq \SI{3}{\kilo\meter}$.  Unlike \citet{klemp2011}, there is no eddy diffusion in the equation set.

The test was integrated forward by 5 hours on the BTF, SLEVE and cut cell grids, and a regular grid with flat terrain.  Maximum vertical velocities are compared with the results from \citet{klemp2011} in figure~\ref{fig:resting} (note different vertical scales).  In agreement with \citet{klemp2011}, vertical velocities are larger on more distorted grids.  However, magnitudes are smaller comparing results on the terrain following grids with those from \citet{klemp2011}, with $w$ reaching a maximum of \(\sim \SI{0.35}{\meter\per\second}\) on the BTF grid in our test compared with a maximum of \(\sim \SI{10}{\meter\per\second}\) found by \citet{klemp2011} \TODO{NO, compare with Klemp's improved results}.

@@

\begin{figure*}
	\centering
	\includegraphics{../msc-publication-fig-resting/fig-resting.pdf}
%
	\caption{Maximum spurious vertical velocity, \(w\) (\si{\meter\per\second}), in the resting atmosphere test with results on (a) BTF, SLEVE, Hybrid Terrain Following (HTF) and Smoothed Terrain Following (STF) coordinates from \citet{klemp2011}, (b) BTF, SLEVE, cut cell and regular grids using the model from \citet{weller-shahrokhi2014}.  Note that vertical scales differ.}
	\label{fig:resting}
\end{figure*}

@@

The initial thermodynamic conditions have a surface temperature of $\theta_0 = \SI{288}{\kelvin}$ and constant stability with $N = \SI{0.01}{\per\second}$ everywhere.  A constant horizontal wind $u = \SI{10}{\meter\per\second}$ is prescribed at the inlet boundary.

@@

No normal flow is imposed at the top and bottom boundaries and the outlet is zero gradient.  For Exner, hydrostatic balance is prescribed on all boundaries.  The simulation is integrated forward by 5 hours with a timestep $\Delta t = \SI{8}{\second}$.

>>>>

The results of the tracer advection test show that numerical errors are due to misalignment of the flow with the grid.  Given an advection scheme of a sufficiently high order, distortions in the grid do not significantly distort the tracer.

@@

At $t = \SI{10000}{\second}$ the tracer has passed over the mountain.  The horizontal position of the tracer centre can be calculated by integrating along the trajectory to find $t$, the time taken to pass from one side of the mountain to the other:
\begin{align}
	\mathrm{d}t &= \mathrm{d}x / u(x) \\
	t &= \int_0^x \frac{H - h(x)}{u_0 H}\:\mathrm{d}x \\
	t &= \frac{x}{u_0} - \frac{h_0}{16 u_0 H} \left[ 4x + \frac{\sin 2 (\alpha + \beta) x}{\alpha + \beta} \right.+ \nonumber \\
   &\ \left. \frac{\sin 2(\alpha - \beta) x}{\alpha - \beta} + 2 \left( \frac{\sin 2\alpha x}{\alpha} + \frac{\sin 2\beta x}{\beta} \right) \right]
\end{align}
Hence, we find that \(x(t=\SI{10000}{\second}) = \SI{51577.4}{\meter}\).  Because the velocity field is non-divergent, the flow accelerates over mountain ridges and the tracer travels \SI{1577.4}{\meter} further compared to advection in the purely horizontal velocity field.  Tracer height is unchanged downwind of the mountains because advection is along the terrain following coordinate surface.

\begin{table*}[t]
	\caption{Minimum and maximum tracer magnitudes and \(\ell^2\) error norms at \(t = \SI{10000}{\second}\) in the horizontal and terrain following tracer advection tests using centred linear and cubic upwind-biased schemes.}
\label{tab:advection}
%
\centering
\footnotesize
\begin{tabular}{ r @{\hspace{0.5em}} l l l | @{\hspace{0.2em}} l l l | @{\hspace{0.5em}} l l l | @{\hspace{0.2em}} l l l}
\hline\hline
& \multicolumn{6}{c}{Horizontal advection} & \multicolumn{6}{c}{Terrain following advection} \\
& \multicolumn{3}{c}{Centred linear} & \multicolumn{3}{c}{Cubic upwind-biased} & \multicolumn{3}{c}{Centred linear} & \multicolumn{3}{c}{Cubic upwind-biased} \\
& \(\ell^2\) error & min & max & \(\ell^2\) error & min & max & \(\ell^2\) error & min & max & \(\ell^2\) error & min & max \\
\hline
Analytic  & 0 & 0 & 1 & 0 & 0 & 1 & 0 & 0 & 1 & 0 & 0 & 1 \\
BTF 	  & \num{0.0318} & \num{-0.442} & \num{0.849} & \num{0.00767} & \num{-0.0433} & \num{0.928} & \num{0.00251} & \num{-0.0244} & \num{0.985} & \num{0.00154} & \num{-0.0105} & \num{0.983} \\
SLEVE 	  & \num{0.00233} & \num{-0.0253} & \num{0.945} & \num{0.00108} & \num{-0.0106} & \num{0.981} & \num{0.0298} & \num{-0.258} & \num{0.865} & \num{0.0122} & \num{-0.0273} & \num{0.864} \\
Cut cell  & \num{0.00224} & \num{-0.0253} & \num{0.985} & \num{0.000579} & \num{-0.00670} & \num{0.982} & \num{0.0318} & \num{-0.270} & \num{0.846} & \num{0.0136} & \num{-0.0297} & \num{0.850} \\
No terrain & \num{0.00224} & \num{-0.0253} & \num{0.985} & \num{0.000577} & \num{-0.00670} & \num{0.982} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & {---} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} \\
\hline
\end{tabular}
\end{table*}

@@

We conclude from this test that accuracy depends upon alignment of the flow with the grid, and accuracy is not significantly reduced by grid distortions.  Error on the BTF grid in the terrain following advection test is comparable with the error on the SLEVE grid in the horizontal advection test.

@@

The initial thermodynamic conditions are in discrete hydrostatic balance, having a reference potential temperature of $\theta(z = 0) = \SI{288}{\kelvin}$ and constant stability with Brunt-V\"ais\"al\"a frequency $N = \SI{0.01}{\per\second}$ everywhere, except for a more stable layer of $N = \SI{0.02}{\per\second}$ between $\SI{2}{\kilo\meter} \leq z \leq \SI{3}{\kilo\meter}$.  Unlike \citet{klemp2011}, there is no eddy diffusion in the equation set.

The test was integrated forward by 5 hours on the BTF, SLEVE and cut cell grids, and a regular grid with flat terrain.  Maximum vertical velocities are compared with the results from \citet{klemp2011} in figure~\ref{fig:resting} (note different vertical scales).  In agreement with \citet{klemp2011}, vertical velocities are larger on more distorted grids.  However, magnitudes are smaller comparing results on the terrain following grids with those from \citet{klemp2011}.  Using the model from \citet{weller-shahrokhi2014}, which includes a curl-free pressure gradient formulation, $w$ reaches a maximum of \(\sim \SI{0.35}{\meter\per\second}\) on the BTF grid, compared with a maximum of \(\sim \SI{7}{\meter\per\second}\) found by \citet{klemp2011} using their improved horizontal pressure gradient formulation.

@@

\begin{figure*}
	\centering
	\includegraphics{../msc-publication-fig-resting/fig-resting.pdf}
%
	\caption{Maximum spurious vertical velocity, \(w\) (\si{\meter\per\second}), in the resting atmosphere test with results on (a) BTF, SLEVE, Hybrid Terrain Following (HTF) and Smoothed Terrain Following (STF) coordinates from \citet{klemp2011} using their improved horizontal pressure gradient formulation, (b) BTF, SLEVE, cut cell and regular grids using the model from \citet{weller-shahrokhi2014} which includes a curl-free pressure gradient formulation.  Note that vertical scales differ.}
	\label{fig:resting}
\end{figure*}

@@

The initial thermodynamic conditions have a surface temperature of $\theta(z=0) = \SI{288}{\kelvin}$ and constant stability with $N = \SI{0.01}{\per\second}$ everywhere.  A constant horizontal wind $u = \SI{10}{\meter\per\second}$ is prescribed at the inlet boundary.

@@

No normal flow is imposed at the top and bottom boundaries and the outlet has a zero gradient boundary condition.  For the Exner function of pressure, hydrostatic balance is prescribed on all boundaries.  The simulation is integrated forward by 5 hours with a timestep $\Delta t = \SI{8}{\second}$.
